Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-02-16 text/paper2/ms.tex
fixed hoggs placeholders

Within the Milky Way our unique vantage point allows us a three-dimensional view of the structure in own dark matter halo. Our proximity allows us to use individual stars as kinematic tracers and hence build much larger samples that probe deeper into the halo than the globular cluster and planetary nebula studies of external galaxies. For example, \cite{deason12a} used halo BHB stars selected from the Sloan Digital Sky Survey \cite[SDSS;][]{york00} as a random tracer population to measure the mass and slope of a simple, power-law fit to the Galactic potential. Such studies assume that the tracer orbits are randomly sampled from a smooth distribution function and phase mixed, however the existence of large-scale substructure in stellar streams and clouds is known to bias mass and velocity inferences that ignore this fact. Another approach is to exploit the non-random nature of the stellar populations in the halo. Tidal streams are dynamically cold systems --- debris typically have small distributions of energy and angular momentum --- and thus require orders of magnitude fewer tracers than a random sample to get constraints of comparable accuracy. For example, in the simplest case we might assume that debris stars are actually still on the same orbit as their progenitor system \citep[a \emph{wrong} assumption, see e.g.,][]{eyre2011}. Then we can imagine measuring the full-space velocities ${\bf v}$ at different points ${\bf x}$ in the structure (e.g., along a stream), which would give us a direct measure of differences in a potential, $\Phi$. 
%In practice, simple methods such as orbit fitting lead to systematic biases in inferred properties of the underlying potential because streams do not strictly delineate orbits [cite eyre binney work].

@@

Although the \gaia\ project will deliver hundreds of millions of stars useful for dynamical inference,
  it still might be that the most relevant stellar samples for inferring the Milky Way potential might be very small in the end.
One reason is that if distances turn out to be important, stars that produce good distance estimates might be much more
  valuable than typical stars in the sample.
At halo distances, \gaia's parallaxes are not expected to be better than the distance indicators produced by the best
  variable stars observed at the best wavelengths; for example, near-infrared observations of RR Lyrae stars
  can in principle deliver XX percent photometric distances at essentially all halo distances (CITATIONS).
These valuable stars are rare (and their abundances are age and metallicity dependent);
  there could be many cold structures in the Milky Way halo that are highly constraining
  on the potential in principle, but which contain only a few good distance-indicating members.
That said, it is not yet known what the trade-offs are between having many stars at low precision and a few at high precision,
  nor is it known how valuable distance information really is, when a structure contains many precisely observed members.

Another reason to think about small samples is more conceptual:
In dynamical inferences that go beyond simple virial or Jeans arguments,
  inference depends on the non-linear dynamical evolution of populations in phase space.
The nonlinearities, combined with heteroskedastic data and requirements to marginalize over phase-space distribution functions,
  lead naturally to situations in which some stars are \emph{far more informative} than others.
For a simple example, in a toy study of the Solar System,
  phase-space distribution marginalization was shown to make some planets are far more informative than others
  about the gravitational potential (CITE Bovy et al).
This is expected to be a generic feature in dynamical problems;
  it might be that a very small fraction of \gaia's catalog entries
  deliver a large fraction of the information about the Milky Way's potential and formation history.

@@

\section{Simulations}\label{sec:sims}
[David H.]
We [...] N-body simulations run with an SCF code [...]. [describe simulations here] [Describe tub]

>>>>

Within the Milky Way our unique vantage point allows us a three-dimensional view of the structure in own dark matter halo. Our proximity allows us to use individual stars as kinematic tracers and hence build much larger samples that probe deeper into the halo than the globular cluster and planetary nebula studies of external galaxies. For example, \cite{deason12a} used halo BHB stars selected from the Sloan Digital Sky Survey \cite[SDSS;][]{york00} as a random tracer population to measure the mass and slope of a simple, power-law fit to the Galactic potential. Such studies assume that the tracer orbits are randomly sampled from a smooth distribution function and phase mixed, however the existence of large-scale substructure in stellar streams and clouds is known to bias mass and velocity inferences that ignore this fact. Another approach is to exploit the non-random nature of the stellar populations in the halo. Tidal streams are dynamically cold systems --- debris typically have small distributions of energy and angular momentum --- and thus require orders of magnitude fewer tracers than a random sample to get constraints of comparable accuracy. For example, in the simplest case we might assume that debris stars are actually still on the same orbit as their progenitor system \citep[a \emph{wrong} assumption, see e.g.,][]{eyre11}. Then we can imagine measuring the full-space velocities ${\bf v}$ at different points ${\bf x}$ in the structure (e.g., along a stream), which would give us a direct measure of differences in a potential, $\Phi$. 
%In practice, simple methods such as orbit fitting lead to systematic biases in inferred properties of the underlying potential because streams do not strictly delineate orbits [cite eyre binney work].

@@

Although the \gaia\ project will deliver hundreds of millions of stars useful for dynamical inference,
  it still might be that the most relevant stellar samples for inferring the Milky Way potential might be very small in the end.
One reason is that if distances turn out to be important, stars that produce good distance estimates might be much more
  valuable than typical stars in the sample.
At halo distances, \gaia's parallaxes are not expected to be better than the distance indicators produced by the best
  variable stars observed at the best wavelengths; for example, near-infrared observations of RR Lyrae stars
  can in principle deliver $\sim$2\% percent photometric distances at essentially all halo distances \citep{madore12}.
These valuable stars are rare (and their abundances are age and metallicity dependent);
  there could be many cold structures in the Milky Way halo that are highly constraining
  on the potential in principle, but which contain only a few good distance-indicating members.
That said, it is not yet known what the trade-offs are between having many stars at low precision and a few at high precision,
  nor is it known how valuable distance information really is, when a structure contains many precisely observed members.

Another reason to think about small samples is more conceptual:
In dynamical inferences that go beyond simple virial or Jeans arguments,
  inference depends on the non-linear dynamical evolution of populations in phase space.
The nonlinearities, combined with heteroskedastic data and requirements to marginalize over phase-space distribution functions,
  lead naturally to situations in which some stars are \emph{far more informative} than others.
For a simple example, in a toy study of the Solar System,
  phase-space distribution marginalization was shown to make some planets are far more informative than others
  about the gravitational potential \citep{bovy10}.
This is expected to be a generic feature in dynamical problems;
  it might be that a very small fraction of \gaia's catalog entries
  deliver a large fraction of the information about the Milky Way's potential and formation history.

@@

\section{Simulations}\label{sec:sims}
[David Hendel]
We [...] N-body simulations run with an SCF code [...]. [describe simulations here] [Describe tub]
